Report on Late Winter2013 Production: Image Differencing

### 2013-03-18 W13report.tex
Added false positives figure, tables with best fits and configs.  ACB

\author{Andrew Becker, Simon Krughoff, Andrew Connolly, Russell Owen}
\title{Summary of Late Winter2013 Production: ip\_diffim}
\date{\today}

@@

We summarize

@@

The focus of this production is to ascertain the ability of the LSST
software stack to provide clean subtractions between a science image
and a reference image (referred to as the template image below).  The
science image will typically be a single--epoch {\tt calexp}, while
the template image will be a high signal--to--noise representation of
the field, typically a {\tt coadd}.

The metric we use to define ``clean'' is the number of false positives
that are detected and measured on the image difference.  We calculate
several additional quality metrics during the image subtraction
process, and correlate them with the number of false positives, as a
means to predict the number of false positives we may expect
downstream from the image subtraction itself.

@@

The basis functions themselves $K_i(u,v)$ are a degree of freedom in
this problem.  The original implementations \citep{Alard98,Alard00}
used a set of 3 Gaussians (represented by our {\tt Config} variable
nGauss), each with a different sigma (sigGauss), and each modified by
a Laguerre polynomial to a given order (degGauss).  Subsequent studies
\citep[e.g.][]{2007AN....328...16I} have suggested that a constant
ratio be maintained between the different Gaussian widths, such that
$\sigma_{i+1} = \beta \times \sigma_{i}$.  We use the value $\beta =
2.0$ for this production.  We normalize the sigmas by noting that,
under the assumption that the Psfs of the images are Gaussian
($\sigma_S$ for the science image and $\sigma_T$ for the template
image), the sigma of the matching kernel should be simply $\sigma_K^2
= \sigma_S^2 - \sigma_T^2$.  We use this width for the central
Gaussian in our sequence, when we use more than one Gaussian, or the
sole Gaussian in the sequence when we use only one.

@@

One complication is that measurement on $D'_{Pre}$ is necessarily
different than on $D'_{Post}$.

@@

Images $S(x,y)$ (optionally pre-filtered with $Psf_S(x,y,u,v)$) and
$T(x,y)$ are sent to ImagePsfMatchTask, whose purpose is to fit for
$K(u,v,x,y)$ and to produce $D(x,y)$.  The first step is to use the
Psf Gaussian sigmas to define the {\it sizes} of the Gaussians shapes
in $K_i(u,v)$.  As mentioned above, the main width is chosen to be
$\sqrt{\sigma_I^2 - \sigma_R^2}$.  We use variations on the
configuration where we use one Gaussian, with the above width, and
three Gaussians, the central of which has this width, and the others a
factor of $\beta = 2.0$ smaller and larger.  We modify each of the
Gaussians by a set of Laguerre polynomials of a given order.  The
smallest Gaussian is modified to the highest order, with the others
modified by order//2.  Figure~\ref{basis} provides an example basis
set with this configuration.  The total number of shapes in the basis
are: $\sum_i^{\rm nGauss} ({\rm degGauss}_i+1)\times({\rm
  degGauss}_i+2)/2$.  The dimensions of the Psf matching kernel are
chosen to be 6 times the largest Gaussian width, with a minimum
kernelSize of 21x21 pixels.  For kernels that are significantly
smaller than this, the Gaussians have significant (non--zero) power at
the kernel boundaries, leading to square systematic artifacts at the
scale of the kernel in the difference images.

@@

For each source $j$ returned by DiaCatalogSourceSelector, we create a
{\tt KernelCandidate} instance, which holds substamps of $T_j(x,y)$
and $S_j(x,y)$ centered on the source.  These will be used to fit for
a local solution $K(u,v)$; the ensemble of local solutions at each
source's $x,y$ will be used to fit for the full spatial solution
$K(u,v,x,y)$.  We emphasize that the dimensions of the substamps is
important.  Note from equation~\ref{eq-soln} that the coefficients
$a_i$ are derived from a convolution of the template substamp with the
kernel basis functions $K_i$.  This convolution means that
kernelSize//2 pixels along each substamp edge are rendered unusable.
To maintain a siginificant number of pixels in $C_i$, we set the
KernelCandidate dimensions to be twice that of the kernel (i.e. the
number of pixels remaining in $C_i$ is equal to the number of pixels
in the kernel, which is fewer than the number of bases $i$).  For each
KernelCandidate, we solve for $K(u,v)$ and create a local difference
image $D_j(x,y)$.  We evaluate several statistics on this local
difference image, normalized by the square root of its variance, which
puts the pixels in units of sigma.  We measure the mean sigma, the RMS
of this distrubion, the $\chi^2$ of these pixels normalized by the
degrees of freedom (number of pixels in $D(x,y)_j$ minus the number of
kernel bases), and the mean squared error.  These are referred to as
the LOCAL kernel metrics, and reflect how appropriate the chosen
kernel basis is.

The ensemble of {\tt KernelCandidates} is then used to constrain a
spatial model of the kernel $K(u,v,x,y)$.  This is done using the {\tt
  SpatialCellSet} formalism, whereby the {\tt KernelCandidates} are
distributed across the image in a grid of {\tt SpatialCells}.  At most
3 {\tt KernelCandidates} per {\tt SpatialCell} are input to a spatial
model of the kernel.  This formalism attempts to distribute
constraints on the spatial model evenly across the image.  For the
spatial kenrel model, we assume that each of the kernel coefficients
$a_i$ may be represented by a $N^{th}$ order 2--dimensional Chebyshev.
Normally, we would perform sigma--clipping iterations on candidates
having bad residuals after spatial modelling.  However in this
production we expect that all objects should fit, so we perform no
sigma clipping.  Having generated the full spatial solution
$K(u,v,x,y) = \sum_i a_i(x,y) K_i(u,v)$, we evaluate the solution at
the position of each {\tt KernelCandidate} and generate the metrics
defined above.  These are referred to as the SPATIAL kernel metrics.
Importantly, we interpolate this solution to the positions of the
control sample, which were explicitly {\it not} used in the spatial
fit, and evaluate these same metrics.  The differences between the
{\tt KernelCandidate} LOCAL and SPATIAL metrics, as well as the values
of the SPATIAL metrics for the control sample, reflect the
appropriateness of the spatial model and our ability to interpolate
and extrapolate to the full extent of the images.

\subsubsection{ImageDifferenceTask}

The main {\tt pipeBase.Task} driving the processing is the
ImageDifferenceTask.  The calls subtasks: DiaCatalogSourceSelector;
RegisterTask; ImagePsfMatchTask; SourceDetectionTask; and
DipoleMeasurementTask.  We describe in detail each subtask below.

\subsubsection{DiaCatalogSourceSelector}

This class queries the reference catalog for {\tt sources} within the
science {\tt exposure's} footprint.  These {\tt sources are used to
  define a set of {\tt KernelCandidates} used to fit the $K(u,v)$,
  each at its specific $x,y$ coordinate.

\subsubsection{RegisterTask}

\subsubsection{ImagePsfMatchTask}

\subsubsection{SourceDetectionTask}

\subsubsection{DipoleMeasurementTask}

@@

@@

We use the ImageDiffereceTask to

@@

\begin{figure}
\includegraphics[width=0.8\textwidth]{figures/shift.eps} \\
\caption{Figure illustrating the ability of the kernel basis to
  compensate, locally, for the effects of misregistration.  Each
  element of the grid shows a single KernelCandidate local difference
  image (visit=6866603, doPreFilter=True).  Columns indicate the
  Gaussian sigma of the basis, which is designed to have one Gaussian
  with the prescribed width, modified to fourth order.  Rows indicate
  the amplitude of the astrometric offset that was introduced into
  the Source positions before image-to-image registration.  
  In the {\it top} row, we see that even with a small astrometric
  offset, the sigma=2 basis is unable to produce a quality
  subtraction, because the width is inappropriate for matching the two
  input Psfs (the theoretical Gaussian sigma here is 3.4 pixels).  At
  sigma=4 the basis is able to produce a quality subtraction, and by
  sigma=5 the basis is too large to match the Psfs.  In {\it row 4},
  we show that even for an astrometric offset of 4 pixels (0.8''), the
  sigma=4 basis can produce a reasonable difference image without
  structured residuals.  However, by {\it row 5}, with an offset of 6
  pixels, none of the Gaussians provide a successful local difference
  image. 
  The ability of the basis to compensate for bulk astrometric offsets
  is a function of the kernel width, and the kernel width is itself a
  function of the two input Psf FWHMs.  Thus there is a complicated
  dependence of the ability of the kernel to shift the centroids of
  stars.  Roughly, the basis may compensate for local astrometric residuals
  at a scale up to $\sqrt{\sigma_I^2 - \sigma_R^2}$.
}
\label{kernel_offsets}
\end{figure}

\bibliographystyle{apj}
\bibliography{refs}

>>>>

\author{Andrew Becker, Simon Krughoff, Andrew Connolly, Russell Owen}
\title{Report on Late Winter2013 Production: Image Differencing}
\date{\today}

@@

We're done!

@@

The focus of this production is to optimize and quantify the false
positive rate in the Psf--matched subtraction of a science image and a
reference image (referred to as the template image below).  The metric
we optimize is the number of false positives that are detected and
measured on the image difference.  We calculate several additional
quality metrics during the image subtraction process, and correlate
them with the number of false positives, as a means to predict the
number of false positives we may expect downstream from the image
subtraction.

@@

The basis functions themselves $K_i(u,v)$ are a degree of freedom in
this problem.  The original implementations \citep{Alard98,Alard00}
used a set of 3 Gaussians (represented by our {\tt Config} variable
nGauss), each with a different sigma (sigGauss), and each modified by
a Laguerre polynomial to a given order (degGauss).  Subsequent studies
\citep[e.g.][]{2007AN....328...16I} have suggested that a constant
ratio be maintained between the different Gaussian widths, such that
$\sigma_{i+1} = \beta \times \sigma_{i}$.  We use the value $\beta =
2.0$ for this production.  We set the scale for sigma by noting that,
under the assumption that the Psfs of the images are Gaussian
($\sigma_S$ for the science image and $\sigma_T$ for the template
image), the sigma of the matching kernel should be simply $\sigma_K^2
= \sigma_S^2 - \sigma_T^2$.  We use this width for the central
Gaussian in our sequence, when we use more than one Gaussian, or the
sole Gaussian in the sequence when we use only one.

@@

%One complication is that measurement on $D'_{Pre}$ is necessarily
%different than on $D'_{Post}$.

@@

Images $S(x,y)$ (optionally pre-filtered with $Psf_S(x,y,u,v)$) and
$T(x,y)$ are sent to ImagePsfMatchTask, whose purpose is to fit for
$K(u,v,x,y)$ and to produce $D(x,y)$.  The first step is to use the
Psf Gaussian sigmas to define the {\it sizes} of the Gaussians shapes
in $K_i(u,v)$.  As mentioned above, the main width is chosen to be
$\sqrt{\sigma_S^2 - \sigma_T^2}$.  We use variations on the
configuration where we use one Gaussian, with the above width, and
three Gaussians, the central of which has this width, and the others a
factor of $\beta = 2.0$ smaller and larger.  We modify each of the
Gaussians by a set of Laguerre polynomials of a given order.  The
smallest Gaussian is modified to the highest order, with the others
modified by order//2.  Figure~\ref{basis} provides an example basis
set with this configuration.  The total number of shapes in the basis
are: $\sum_i^{\rm nGauss} ({\rm degGauss}_i+1)\times({\rm
  degGauss}_i+2)/2$.  The dimensions of the Psf matching kernel are
chosen to be 6 times the largest Gaussian width, with a minimum
kernelSize of 21x21 pixels.  For kernels that are significantly
smaller than this, the Gaussians have significant (non--zero) power at
the kernel boundaries, leading to square systematic artifacts at the
scale of the kernel in the difference images.

@@

For each source $j$ returned by DiaCatalogSourceSelector, we create a
{\tt KernelCandidate} instance, which holds substamps of $T_j(x,y)$
and $S_j(x,y)$ centered on the source.  These will be used to fit for
a local solution $K(u,v)$; the ensemble of local solutions at each
source's $x,y$ will be used to fit for the full spatial solution
$K(u,v,x,y)$.  We emphasize that the dimensions of the substamps is
important.  Note from equation~\ref{eq-soln} that the coefficients
$a_i$ are derived from a convolution of the template substamp with the
kernel basis functions $K_i$.  This convolution means that
kernelSize//2 pixels along each substamp edge are rendered unusable.
To maintain a significant number of pixels in $C_i$, we set the
KernelCandidate dimensions to be twice that of the kernel (i.e. the
number of pixels remaining in $C_i$ is equal to the number of pixels
in the kernel, which is fewer than the number of bases $i$).  For each
KernelCandidate, we solve for $K(u,v)$ and create a local difference
image $D_j(x,y)$.  We evaluate several statistics on this local
difference image, normalized by the square root of its variance, which
puts the pixels in units of sigma.  We measure the mean sigma, the RMS
of this distribution, the $\chi^2$ of these pixels normalized by the
degrees of freedom (number of pixels in $D(x,y)_j$ minus the number of
kernel bases), and the mean squared error.  These are referred to as
the LOCAL kernel metrics, and reflect how appropriate the chosen
kernel basis is.

The ensemble of {\tt KernelCandidates} is then used to constrain a
spatial model of the kernel $K(u,v,x,y)$.  This is done using the {\tt
  SpatialCellSet} formalism, whereby the {\tt KernelCandidates} are
distributed across the image in a grid of {\tt SpatialCells}.  At most
3 {\tt KernelCandidates} per {\tt SpatialCell} are input to a spatial
model of the kernel.  This formalism attempts to distribute
constraints on the spatial model evenly across the image.  For the
spatial kernel model, we assume that each of the kernel coefficients
$a_i$ may be represented by a $N^{th}$ order 2--dimensional Chebyshev.
Normally, we would perform sigma--clipping iterations on candidates
having bad residuals after spatial modeling.  However in this
production we expect that all objects should fit, so we perform no
sigma clipping.  Having generated the full spatial solution
$K(u,v,x,y) = \sum_i a_i(x,y) K_i(u,v)$, we evaluate the kernel
solution at the position of each {\tt KernelCandidate}, create a
difference image using this kernel, and recalculate the metrics
defined above.  These are referred to as the SPATIAL kernel metrics.
Importantly, we also interpolate this solution to the positions of the
control sample, which were explicitly {\it not} used in the spatial
fit, and evaluate these same metrics.  The differences between the
{\tt KernelCandidate} LOCAL and SPATIAL metrics, as well as the values
of the SPATIAL metrics for the control sample, reflect the
appropriateness of the spatial model and our ability to interpolate
and extrapolate to the full extent of the images.  We use the Mean
Square Error (MSE) of the control sample to example the tradeoff
between bias and variance in the spatial interpolation.

We next run the SourceDetectionTask on the difference image, using a
``polarity'' of {\it both} to search for both positive and negative
deviations.  We use a nominal detection threshold of 5 sigma.  In the
case of pre--filtering, we operate on the difference image directly.
In post--filtering, we correlate the difference image with its Psf (by
design, the same as the Psf of $S$) before detection is run.  All else
being equal, the images immediately preceding the detection step
($D'$) should be {\it exactly} the same.  We expect minor differences
in the variance due to the fact that the post--filtered data will have
gone through 3 convolutions (astrometric resampling; Psf--matching to
the science image; Psf filtering for detection) while the
pre--filtered data will have gone through only two (astrometric
resampling; psf--matching to the maximum likelihood image).
Differences in the shapes and sizes of the Psf--matching kernels used
in the two cases may also cause systematic differences in the images.
Detections are merged using a grow radius of 2 pixels.

Finally, we run DipoleMeasurementTask on the detections to create {\tt
  DiaSources}.  DipoleMeasurementTask is an extension of
SourceMeasurementTask that includes specific dipole measurement
algorithms.  Measurement is necessarily different in the pre--filtered
and post--filtered data.  In particular, in post--filtering, detection
happens on the filtered maximum likelihood image and source
measurement on the associated pixels in the unfiltered difference
image.  In pre--filtering, the latter does not exist, and thus
detection and measurement happen on the same image.  This restricts
the types of measurements that may be made on the sources.  In
particular, the Psf flux is merely the fitted amplitude of the peak in
the image, and the Psf centroid its center.  While these measurements
are trivial to implement, it is unclear how to implement the remainder
of the measurement suite in manner comparable to the measurements on
post--filtered data.

All {\tt DiaSources} are associated with the {\tt calexp's} {\tt
  sources} and with the reference catalog, with a permissive matching
radius of 2''.  {\tt DiaSources} that match with {\tt sources} may be
true residuals around stars {\it or} true false positives that are
also found in {\tt calexp} SourceDetectionTask.  For this reason, we
use the associations with the reference catalog objects as the
assessment if a given {\tt DiaSource} matches with a {\tt source} or
whether it is an orphan (e.g. a statistical fluctuation in the
background).  In addition, all {\tt KernelCandidates} are persisted
with the above {\tt LOCAL} and {\tt SPATIAL} metrics.  Repositories
are accessed with a specially written RepositoryIterator that
aggregates {\tt KernelCandidates} from a given {\tt source} across
multiple output repositories.

@@

We ran several pre--productions to tune: the dimensions of the kernels
and KernelCandidates; the widths of the Gaussians (sigGauss) based on
the FWHM of the input images; and the scaling of the Gaussians
($\beta$) in the basis sequence.  The main production runs included:
an outer loop over the use of pre--filtering or post--filtering for
detection; variations of the complexity of the Kernel basis, including
the number of Gaussians (nGauss = 1,3) and the degree of the modifying
Laguerre polynomials (degGauss = 1..6); and the spatial order of the
Chebyshev polynomials on $a_i(x,y)$ (spatialKernelOrder = 1..6).
Finally, we ran several post--productions focused on perturbing the
solutions that yielded the fewest numbers of DiaSources along specific
dimensions.  This included: the RMS of the WCS solution used by
RegisterTask; including a DC offset in the source positions prior to
RegisterTask, to determine the ability of the kernel basis to
accommodate astrometric misregistration; the size of the SpatialCells
used in the spatial interpolation, which effectively regulates the
numbers of constraints that go into the spatial fit; and modifying the
detection thresholds to determine how the number of false positives
scales with threshold.

@@

These algorithms were run on 3 visits, with seeings 0.6'' (visit
6866601), 0.88'' (visit 6866602), and 1.2'' (visit 6866603).  These
visits were simulated with {\bf DETAIL}.  All 9 sensors of the central
raft were simulated, for a total of 27 difference images per
configuration to use in our comparative analysis.

A template image was generated by {\bf DETAIL}.

@@

\section{Conclusions}

We can reach near the theoretical limit for false positive rate in
these ideal data, something which has not been demonstrated for image
subtraction algorithms in the past.

Pre--filtering is clearly preferred when the traditional process would
lead to deconvolution, or sharpening, of the data.  When the standard
processing results in a convolution, or smoothing of the data,
pre--filtering provides fewer false positives.  As a trade--off, the
measurement of these false positives is not yet well defined.

\section{Future Work}

We intend to research additional aspects of the problem going forward,
including:
\begin{itemize}
\item How the false positive rate varies with more realistic SEDs;
\item How the false positive rate varies with airmass;
\item How the system behaves with a realistic mix of stars and galaxies;
\item How the false positive rate looks in crowded fields;
\item How to perform measurement on pre--filtered difference images;
\end{itemize}

\clearpage
\bibliographystyle{apj}
\bibliography{refs}

\clearpage
\begin{table}
\centering
\begin{tabular}{clcc}
\hline
\multicolumn{4}{|c|}{Best Results} \\
\hline
Visit    & Prefilter & False Positive Matches & False Positive Orphans \\
\hline
v6866601 & True      & 5                      & 29  \\
         & False     & 10                     & 81  \\
v6866602 & True      & 5                      & 30  \\
         & False     & 9                      & 35  \\
v6866603 & True      & 2                      & 9  \\
         & False     & 1                      & 23  \\
\end{tabular}
\caption{Total number of {\tt DiaSources} from all 9 sensors in raft
  2,2, for the configurations that yielded the lowest number of false
  positives.  Matches are determiend using a 2'' match radius with the
  input reference catalog.  Objects with the {\tt EDGE} flag are
  ignored. \label{tab-bestfp}}
\end{table}

\begin{table}
\centering
\begin{tabular}{clccc}
\hline
\multicolumn{5}{|c|}{Best Configs} \\
\hline
Visit    & Prefilter & nGauss & degGauss & spatialOrder \\
\hline
v6866601 & True      & 3      & 6        & 5 \\
         & False     & 3      & 2        & 4 \\
v6866602 & True      & 3      & 6        & 5 \\
         & False     & 3      & 5        & 5 \\
v6866603 & True      & 3      & 6        & 4 \\
         & False     & 3      & 6        & 4 \\
\end{tabular}
\caption{Configuations that led to the best results listed in
  Table~\ref{tab-bestfp}.  All post-production runs include
  perturbations about these configurations. \label{tab-bestconfig}}
\end{table}

\clearpage

\begin{figure}
\includegraphics[width=1.0\textwidth]{figures/falsePos.eps} \\
\caption{``Heat--map'' detailing the total numbers of false positives
  as a function of two main degrees of freedom: the maximum order of
  basis Laguerre polynomials along the y--axis, and the spatial order
  of the Chebyshev expansion of the spatial model along the x--axis.
  The {\it left} column shows these results for prefiltering, while
  the {\it right} column shows this for the post--filtered analysis.
  The best results listed in Table~\ref{tab-bestfp}, with the
  associated configureations from Table~\ref{tab-bestconfig},
  correspond to the minimum in each of the subpanels.
}
\label{fp_heatmap}
\end{figure}

\begin{figure}
\includegraphics[width=0.8\textwidth]{figures/shift.eps} \\
\caption{Figure illustrating the ability of the kernel basis to
  compensate, locally, for the effects of misregistration.  Each
  element of the grid shows a single KernelCandidate local difference
  image (visit=6866603, doPreFilter=True).  Columns indicate the
  Gaussian sigma of the basis, which is designed to have one Gaussian
  with the prescribed width, modified to fourth order.  Rows indicate
  the amplitude of the astrometric offset that was introduced into
  the Source positions before image-to-image registration.  
  In the {\it top} row, we see that even with a small astrometric
  offset, the sigma=2 basis is unable to produce a quality
  subtraction, because the width is inappropriate for matching the two
  input Psfs (the theoretical Gaussian sigma here is 3.4 pixels).  At
  sigma=4 the basis is able to produce a quality subtraction, and by
  sigma=5 the basis is too large to match the Psfs.  In {\it row 4},
  we show that even for an astrometric offset of 4 pixels (0.8''), the
  sigma=4 basis can produce a reasonable difference image without
  structured residuals.  However, by {\it row 5}, with an offset of 6
  pixels, none of the Gaussians provide a successful local difference
  image. 
  The ability of the basis to compensate for bulk astrometric offsets
  is a function of the kernel width, and the kernel width is itself a
  function of the two input Psf FWHMs.  Thus there is a complicated
  dependence of the ability of the kernel to shift the centroids of
  stars.  Roughly, the basis may compensate for local astrometric residuals
  at a scale up to $\sqrt{\sigma_S^2 - \sigma_T^2}$.
}
\label{kernel_offsets}
\end{figure}
